---
title: "p8105_hw3_iah2124"
author: "Iris Hart"
date: "2024-10-10"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Problem 1

Loading packages + data
```{r}
library(tidyverse)

library(p8105.datasets)
data("ny_noaa")
```
EDIT STILL
There are seven variables and 2,595,176 observations in the NY NOAA data set. The variables are ID, date, prcp, snow, snwd, tmax, and tmin. 

Create year, month, and day variables
```{r}
library(dplyr)
library(lubridate)

ny_noaa_clean = ny_noaa %>%
  mutate(
    year = year(date),
    month = month(date),
    day = day(date)
  )
```
View to make sure new variables are created
```{r}
view(ny_noaa_clean)
```

Convert precipitation variable to be mm instead of tenths of mm. Convert daily temperature min and max variables to be degrees to be degrees Celsius from tenths of degrees Celsius. 
```{r}
ny_noaa_clean = ny_noaa %>%
  mutate(
    tmax = as.numeric(tmax) / 10,   
    tmin = as.numeric(tmin) / 10,    
    prcp = as.numeric(prcp) / 10     
  )
```
Check to see if changes were made
```{r}
view(ny_noaa_clean)
```

